biological data preperation

### 2019-01-14 src/biological_data_prep.Rmd
minor edit to how gear == NA reported (changed to unknown)
investingating presence data for missing values

```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap", "dplyr","plyr", "rgdal", "sp", "raster", "tibble", "useful", "marmap", "weathermetrics")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

@@

@@

>>>>

```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap", "dplyr","plyr", "rgdal", "sp", "raster", "tibble", "useful", "marmap", "weathermetrics", "fastmatch")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

@@

#gear type adjustment

should not have gear not reported listed as NA (for some analysis). Change gear == NA to unknown

```{r}
data_aea$gear <- as.character(data_aea$gear)
data_aea$gear[is.na(data_aea$gea)] <- "unknown"
write.csv(data_aea, "../output/bio/data_aea_cell_amo_nao_bins_env_depth_cellobno_env_naocor_cels.csv", row.names = FALSE)
```

@@

# checking data for NA's

The presence points have a lot of NAs in the env. correlates. I want to:

-ID and remove rows where all values == NA
-ID observations with no depth data that are missing surface values (will need to be removed)
-ID observations with depth data but are missing some surface and/or depth values

how many rows(observations do you have)
```{r}
nrow(data_aea) # there are 11516 rows
```

check for no rows (observations) with NAs

```{r}
presence_noNAcount <- sum(!complete.cases(data_aea)) #this just counts the number of rows that contain NA values
presence_noNAcount
```
I have 4687 observations that contain NAs

1) find ones that have no env. data at all. these will need to be removed straight off.

There are 11516 rows in the original presence-point dataset

find the column numbers that correspond to the env. data (use fastmatch package as is quicker)

```{r}
fmatch("chl_surface", names(data_aea)) #28
fmatch("chl_depth", names(data_aea)) #29
fmatch("mlp_surface", names(data_aea)) #30
fmatch("o2_surface", names(data_aea)) #31
fmatch("o2_depth", names(data_aea)) #32
fmatch("salinity_surface", names(data_aea)) #33
fmatch("salinity_depth", names(data_aea)) #34
fmatch("ssh_surface", names(data_aea)) #35
fmatch("temp_surface", names(data_aea)) #36
fmatch("temp_depth", names(data_aea)) #37
```
ok so the env. data i am intersted in is in cols 28:37

Out of curiosoity, map the location of these points. Note the difference between this and the code below is the addition of ! on rowSums

```{r}
presence_novals <- data_aea[!rowSums(is.na(data_aea[ , 28:37])) !=10, ] #4:13 represent the columns w/ env. data, 10 is the number of columns to check (4 to 13 inclusive)
write.csv(presence_novals, "../output/bio/presence_points_without_envdata.csv", row.names = FALSE)
nrow(presence_novals) #just to check it has mapped
```

So I had 123 rows without any env. data (11516 - 11393).

plot
```{r map presences without any env. data}
map <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map, xlim = c(-70, -43), ylim =c(38,70), asp = 1) #the x and y lim are the long-lat bounds of the map
points(presence_novals$decimalLongitude, presence_novals$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
```

Ok so they are along the coast, so the points probably fell outside the boxes.

Now remove the values
```{r}
data_aea <- data_aea[rowSums(is.na(data_aea[ , 28:37])) !=10, ] #4:13 represent the columns w/ env. data, 10 is the number of columns to check (4 to 13 inclusive)
write.csv(data_aea, "../output/bio/presence_nodata_rm")
nrow(data_aea)
```

ok so the next step is to make sure that any of the data you have without a depth value does have all relevant surface values

maybe subset

```{r}
presence_nodepth = data_aea[is.na(data_aea$depth), ]
nrow(presence_nodepth) #1215 rows
head(presence_nodepth)
```

go through each value and see which are missing data
```{r}
presence_nodepth_nochl <- presence_nodepth[is.na(presence_nodepth$chl_surface), ]
nrow(presence_nodepth_nochl) #just to check it has mapped
presence_nodepth_noo2 <- presence_nodepth[is.na(presence_nodepth$o2_surface), ]
nrow(presence_nodepth_noo2) #just to check it has mapped
presence_nodepth_nosalinity <- presence_nodepth[is.na(presence_nodepth$salinity_surface), ]
nrow(presence_nodepth_nosalinity) #just to check it has mapped
presence_nodepth_notemp <- presence_nodepth[is.na(presence_nodepth$temp_surface), ]
nrow(presence_nodepth_notemp) #just to check it has mapped
presence_nodepth_nossh <- presence_nodepth[is.na(presence_nodepth$ssh_surface), ]
nrow(presence_nodepth_nossh) #just to check it has mapped
presence_nodepth_nomlp <- presence_nodepth[is.na(presence_nodepth$mlp_surface), ]
nrow(presence_nodepth_nomlp) #just to check it has mapped
```

great! all of the values that have no depth data have surface data.

now on to the data which has depth data

```{r}
presence_wdepth = data_aea[!is.na(data_aea$depth), ]
nrow(presence_wdepth) #10178 rows
head(presence_wdepth)
```

go through each value and see which are missing data. First the surface values
```{r}
presence_wdepth_nochl <- presence_wdepth[is.na(presence_wdepth$chl_surface), ]
nrow(presence_wdepth_nochl) #just to check it has mapped
presence_wdepth_noo2 <- presence_wdepth[is.na(presence_wdepth$o2_surface), ]
nrow(presence_wdepth_noo2) #just to check it has mapped
presence_wdepth_nosalinity <- presence_wdepth[is.na(presence_wdepth$salinity_surface), ]
nrow(presence_wdepth_nosalinity) #just to check it has mapped
presence_wdepth_notemp <- presence_wdepth[is.na(presence_wdepth$temp_surface), ]
nrow(presence_wdepth_notemp) #just to check it has mapped
presence_wdepth_nossh <- presence_wdepth[is.na(presence_wdepth$ssh_surface), ]
nrow(presence_wdepth_nossh) #just to check it has mapped
presence_wdepth_nomlp <- presence_wdepth[is.na(presence_wdepth$mlp_surface), ]
nrow(presence_wdepth_nomlp) #just to check it has mapped
```

Great! All have surface values. now to the depth values

```{r}
presence_wdepth_nochl <- presence_wdepth[is.na(presence_wdepth$chl_depth), ]
nrow(presence_wdepth_nochl) #just to check it has mapped
presence_wdepth_noo2 <- presence_wdepth[is.na(presence_wdepth$o2_depth), ]
nrow(presence_wdepth_noo2) #just to check it has mapped
presence_wdepth_nosalinity <- presence_wdepth[is.na(presence_wdepth$salinity_depth), ]
nrow(presence_wdepth_nosalinity) #just to check it has mapped
presence_wdepth_notemp <- presence_wdepth[is.na(presence_wdepth$temp_depth), ]
nrow(presence_wdepth_notemp) #just to check it has mapped
```

damn... so

chl_depth = 2394 missing values
o2_depth = 2394 missing values
salinity_depth = 3133 missing values
temp_depth = 3109 missing values
